Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-08-04 manu_draft.Rmd
maybe renaming gams

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19], including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.  This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.  Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.  We present four GAM structures, following a similar format as @Murphy19.  The notation for gam3, gam4, and gam5 are not used because these describe models in @Murphy19 that were developed for purposes specific to Chesapeake Bay.

\begin{equation}
\boldsymbol{gam1:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
(\#eq:gamone)
\end{equation}

\begin{equation}
\boldsymbol{gam2:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
(\#eq:gamtwo)
\end{equation}

\begin{equation}
\boldsymbol{gam6:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2^*\left(year\right) + \epsilon
(\#eq:gamsix)
\end{equation}

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.  Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.

@@

Each of the four GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.  For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoothers terms.  Models were also evaluated using different transformations to determine which had improved fit for specific time series.  Secondary analyses evaluating trends with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.  Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.

@@

```{r, results = 'hide'}
# get predictions for all four gams
tomod <- rawdat %>%
  filter(station %in% 22) %>%
  filter(param %in% 'chl')

@@

png('figs/modsumfig.png', height = 2.5, width = 6, units = 'in', family = 'serif', res = 400)
print(p1)
dev.off()
```
```{r modsumfig, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).  Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
knitr::include_graphics('figs/modsumfig.png')
```

@@

```{r modprfboxtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using BoxCox transformations. Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.  Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized. Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- modprf %>% 
  filter(trans != 'log10') %>% 
  select(-trans) %>% 
  group_by(station) %>% 
  mutate(
    AIC = AIC - max(AIC),
    minaic = AIC == min(AIC), 
    mingcv = GCV == min(GCV),
    maxr2 = R2 == max(R2)
  ) %>% 
  ungroup

@@

```{r modprflogtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using logarithmic (base-10) ransformations. Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.  Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized. Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- modprf %>% 
  filter(trans == 'log10') %>% 
  select(-trans) %>% 
  group_by(station) %>% 
  mutate(
    AIC = AIC - max(AIC),
    minaic = AIC == min(AIC), 
    mingcv = GCV == min(GCV),
    maxr2 = R2 == max(R2)
  ) %>% 
  ungroup

>>>>

To evaluate GAMs with SFE chlorophyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19].  We specifically build our models using the `gam1` and `gam2` structures in @Murphy19, but present a different naming convention to better distinguish the functional components of the models.  The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Additive" and "Interactive", respectively.  Below, we describe the "Additive*" star model as a slight modification of the former. Each model has a similar functional form for time in relation to chlorophyll with slight differences in the smoothers that describe time.

\begin{equation}
\boldsymbol{Additive:} g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
(\#eq:gamone)
\end{equation}

\begin{equation}
\boldsymbol{Interactive:} g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
(\#eq:gamtwo)
\end{equation}

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The three models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.  Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.

@@

Each of the three GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.  For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoothers terms.  Models were also evaluated using different transformations to determine which had improved fit for specific time series.  Secondary analyses evaluating trends with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.  Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.

@@

```{r, results = 'hide'}
# get predictions for all three gams
tomod <- rawdat %>%
  filter(station %in% 22) %>%
  filter(param %in% 'chl')

@@

png('figs/modsumfig.png', height = 2.5, width = 6, units = 'in', family = 'serif', res = 400)
print(p1)
dev.off()
```
```{r modsumfig, fig.cap = "General output for each of the three GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).  Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
knitr::include_graphics('figs/modsumfig.png')
```

@@

```{r modprfboxtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using BoxCox transformations. Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.  Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized. Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- modprf %>% 
  filter(trans != 'log10') %>% 
  select(-trans) %>% 
  group_by(station) %>% 
  mutate(
    AIC = AIC - max(AIC),
    minaic = AIC == min(AIC), 
    mingcv = GCV == min(GCV),
    maxr2 = R2 == max(R2)
  ) %>% 
  ungroup

@@

```{r modprflogtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using logarithmic (base-10) ransformations. Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.  Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized. Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- modprf %>% 
  filter(trans == 'log10') %>% 
  select(-trans) %>% 
  group_by(station) %>% 
  mutate(
    AIC = AIC - max(AIC),
    minaic = AIC == min(AIC), 
    mingcv = GCV == min(GCV),
    maxr2 = R2 == max(R2)
  ) %>% 
  ungroup
